---
editor: 
  markdown: 
    wrap: 72
---

## Learning Objectives {.unnumbered}

-   Practice base R skills
-   Practice Markdown syntax
-   Work in Quarto document

## Introduction

In this session of R practice, we will be working with the dataset:
[Tobias Schwoerer, Kevin Berry, and Jorene Joe. 2022. A household survey
documenting experiences with coastal hazards in a western Alaska
community (2021-2022). Arctic Data Center.
doi:10.18739/A29Z90D3V.](https://doi.org/10.18739/A29Z90D3V).

We will be focusing on the initial survey results
(Initial_Survey111721_ADC.csv). In this file, individual survey
responses are oriented as rows, and the questions are oriented as
columns. The column names are Q1, Q2, etc. Information about what was
the asked question, and what the allowed values mean, are available in
the metadata for each file. **You can access the metadata for each file
by clicking the "more info" link next to the file name at the top of the
page.**

::: callout-note
## Big Idea

The goal for this session is to practice downloading data, reading it
into R in a Quarto document, using base R commands to summarize a
variable within the dataset, and practice formatting a Quarto document
using Markdown syntax.
:::

## Exercise 1

::: callout-tip
## Set up

-   Navigate to the dataset archived in the [Artic Data Center
    repository](https://doi.org/10.18739/A29Z90D3V).
-   Download the file `Initial_Survey111721_ADC.csv`
-   Move the file to a `data` folder within your training project in the
    server.
    -   For this you have to click on the *Upload* button on the *Files
        Pane* in RStudio.
    -   A pop-up window will come up. Make sure that the *Target
        directory* (directory where you file is going to be saved to) is
        the `data` folder withing your RProject.
    -    Click on *Choose File* and navigate to where
        `Initial_Survey111721_ADC.csv` was downloaded in you local
        computer. Click *Open*, then press *OK*.

You should now see the `Initial_Survey111721_ADC.csv` inside the `data`
folder in your project.
:::

### Open a new Quarto Document

::: callout-note
## Step 1

Create a new Quarto document, and structure your documents with the
following headers:

-   Read data

-   Explore data

-   Calculate mean

-   Conclusion

**Tip:** In Markdown syntax, `#` indicates the level of header. `#` is a
“level one header”, meaning the biggest font and the top of the
hierarchy. `###` is a level three header, and will show up nested below
the `#` and `##` headers.
:::

### Read in the data

::: callout-note
## Step 2

Read in the data and store the data frame as `survey_data`.
:::

::: {.callout-note appearance="minimal" collapse="true"}
## Example Code

```{r}
survey_data <- read.csv("data/Initial_Survey111721_ADC.csv")
```

:::

### Explore the data

::: callout-note
## Step 3

Try functions like `summary()`, `colnames()`, `str()`, `unique()`. Feel
free to use any other function you'd like to learn and explore this data
set.

You can also try `View()` in the console. Note that if you include
`View()` in your Quarto document it will cause an error when rendering.
:::

::: {.callout-note appearance="minimal" collapse="true"}
## Example Code

```{r}
summary(survey_data)
colnames(survey_data)
str(survey_data)
unique(survey_data$Q6_2)

```

:::

### Calculate the mean

::: callout-note
## Step 4

Calculate the mean of the answers to Question 3 and save this value into
an object in your Global Environment.

**Tip:** Look at the help page if your answer isn't what you expect,
`?mean()`. Does Q3 column have `NA` values?
:::


::: {.callout-note appearance="minimal" collapse="true"}
## Example Code

```{r}

q3_mean <- mean(survey_data$Q3, na.rm = TRUE)

```


:::


### Write a conclusion

::: callout-note
## Step 5

Interpret this value using the metadata for the table and write a
conclusion based on your interpretation calling the mean value you
calculated in text.
:::

::: {.callout-note appearance="minimal" collapse="true"}
## Example
_Your Markdown file should look something like this_

The average score about respondents perspective in xyz is \`r `round(`q3_mean)`\`. 
This means that ...

:::

### Render and Code Chunk Options

::: callout-note
## Step 6

Render your Quarto file and make sure you get the output you are expecting. Are you running into any errors? If your file is not rendering as you expect discuss with your neighbor how you can improve your file.

Did you added any code chunk options? Review Section [A Quarto Document](https://learning.nceas.ucsb.edu/2024-01-arctic/session_06.html#a-quarto-document) on the Literate Analysis Lesson and decide what code chunck otion you want to include in your file.
:::


## Bonus

::: callout-note
## Go Further
What other ways might you summarize the answers to question 3? Explore!
:::