## Hands On: Clean and Integrate Datasets

### Learning Objectives

In this lesson, you will:

-   Clean and integrate a dataset using dplyr and tidyr
-   Make use of previously-learned knowledge of dplyr, tidyr, and writing functions

### Outline

In this session, you will load data from the following dataset:

[Richard Lanctot and Sarah Saalfeld. 2019. Utqiaġvik shorebird breeding ecology study, Utqiaġvik, Alaska, 2003-2018. Arctic Data Center. doi:10.18739/A23R0PT35](https://doi.org/10.18739/A23R0PT35)

Using our knowledge of R, we are going to try to answer the following two questions:

-   What species of shorebird has the largest eggs and has this changed through time?
-   What species of predator is the most abundant and has this changed through time?

One of the features if this dataset is that it has many files with similar formatting, most of which contain the column `species` which is comprised of the [Bird Banding Laboratory species codes](https://www.pwrc.usgs.gov/bbl/manual/speclist.cfm). These four letter codes aren't very intuitive to most people, so one of the things we will do is write a function that can be used on any file in this dataset that contains a species code.

#### High-level steps {.unnumbered}

The goal here is for you to have to come up with the functions to do the analysis with minimal guidance. This is supposed to be challenging if you are new to dplyr/tidyr. Below is a set of high-level steps you can follow to answer our research question. After the list is a schematic of the steps in table form which I expect will be useful in guiding your code.

1.  Load the species table using the code in the Setup block below

2.  Read the following two files into your environment

    -   Utqiagvik\_egg\_measurements.csv
    -   Utqiagvik\_predator\_surveys.csv

3.  Write a function that will translate species codes into common names.

    -   **Hint**: The fastest way to do this involves adding a column to the `data.frame`. Your function will have two arguments
    -   **Optional Extra Challenge**: For a little extra challenge, try to incorporate an `if` statement that looks for `NA` values in the common name field you are adding. What other conditionals might you include to make your function smarter?

4.  Calculate mean egg volume by year and species

    -   for simplicity, use the formula that assumes the egg is a symmetrical ellipsioid: $V = 4/3{\pi}L(W/2)^2$

5.  Plot mean egg volume by year and species

6.  Calculate total number of predators by year and species

7.  Plot total predators by year and species

#### Setup {.unnumbered .setup}

Use the following two code chunks to set up your species table. The code I use here incorporates two new packages that I thought would be worth showing.

The first, `rvest`, is a package that enables easy scraping and handling of information from websites. It requires a moderate amount of knowledge of `html` to use, but can be very handy. In this case, I use it because I couldn't find a plain text version of the BBL species codes anywhere. To build a reproducible and long lived workflow, I would want to run this code and then store a plain text version of the data in a long lived location, which cites the original source appropriately.

The second package, `janitor`, is a great package that has simple tools for cleaning up data. It is especially handy for handling improperly named variables in a `data.frame.` In this case, I use it to handle a bunch of spaces in the column names so that our lives are a little easier later on.

```{r}
library(rvest)
library(readr)
library(dplyr)
library(janitor)
```

```{r}
webpage <- read_html("https://www.pwrc.usgs.gov/bbl/manual/speclist.cfm")

tbls <- html_nodes(webpage, "table") %>% 
    html_table(fill = TRUE)

species <- tbls[[1]] %>% 
    clean_names() %>% 
    select(alpha_code, common_name) %>% 
    mutate(alpha_code = tolower(alpha_code))
```

#### Visual schematic of data {-}

Make this:

    year common_name            mean_volume
    2003 American Golden-Plover      60426
    2003 Dunlin                      24702
    2003 Long-billed Dowitcher       43824
    2003 Pectoral Sandpiper          27369

and then make this:

    year common_name        total_count
    2003 Glaucous Gull               54
    2003 Parasitic Jaeger             2
    2003 Pomarine Jaeger              6
    2004 Glaucous Gull               69
    2004 Long-tailed Jaeger          13
    

#### Aside {- .aside}

**Why do we need to use a function for this task?**

You will likely at some point realize that the function we asked you to write is pretty simple. The code can in fact be accomplished in a single line. So why write your own function for this? There are a couple of answers. The first and most obvious is that we want to you practice writing function syntax with simple examples. But there are other reasons why this operation might benefit from a function:

* Follow the DRY principles!
    - If you find yourself doing the same cleaning steps on many of your data files, those operations are good candidates for functions. This falls into that category, since we need to do the same transformation on both of the files we use here, and if we incorporated more files from this dataset it would come in even more use.
* Add custom warnings and quality control.
    - Functions allow you to incorporate quality control through conditional statements coupled with warnings. Instead of checking for NA's or duplicated rows after you run a join, you can check within the function and return a warning if any are found.
* Check your function input more carefully
    - Similar to custom warnings, functions allow you to create custom errors too. Writing functions is a good way to incorporate defensive coding practices, where potential issues are looked for and the process is stopped if they are found.

### Full solution. Warning, spoilers ahead!



```{r, warning = FALSE, message = FALSE}
library(rvest)
library(readr)
library(dplyr)
library(janitor)
library(ggplot2)
```

```{r}
webpage <- read_html("https://www.pwrc.usgs.gov/bbl/manual/speclist.cfm")

tbls <- html_nodes(webpage, "table") %>% 
    html_table(fill = TRUE)

species <- tbls[[1]] %>% 
    clean_names() %>% 
    select(alpha_code, common_name) %>% 
    mutate(alpha_code = tolower(alpha_code))
```

```{r, warning=F, message=F}
egg <- read.csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A4b219711-2282-420a-b1d6-1893fe4a74a6")

pred <- read.csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A9ffec04c-7e2d-41dd-9e88-b6c2e8c4375e")
```

```{r}
#' Function to add common name to data.frame according to the BBL list of species codes

#' @param df A data frame containing BBL species codes in column `species`
#' @param species A data frame defining BBL species codes with columns `alpha_code` and `common_name`
#' @return A data frame with original data df, plus the common name of species

assign_species_name <- function(df, species){
    if (!("alpha_code" %in% names(species)) |
        !("species" %in% names(df)) |
        !("common_name" %in% names(species))){
      stop("Tables appear to be formatted incorrectly.")
    }  
  
    return_df <- left_join(df, species, by = c("species" = "alpha_code"))
    
    if (nrow(return_df) > nrow(df)){
      warning("Joined table has more rows than original table. Check species table for duplicated code values.")
    }
    
    if (length(which(is.na(return_df$common_name))) > 0){
      x <- length(which(is.na(return_df$common_name)))
      warning(paste("Common name has", x, "rows containing NA"))
    }
    
    return(return_df)
        
}
```

Or a simple version without the extra challenges added:

```{r, eval = F}
#' Function to add common name to data.frame according to the BBL list of species codes

#' @param df A data frame containing BBL species codes in column `species`
#' @param species A data frame defining BBL species codes with columns `alpha_code` and `common_name`
#' @return A data frame with original data df, plus the common name of species

assign_species_name <- function(df, species){
    return_df <- left_join(df, species, by = c("species" = "alpha_code"))
    return(return_df)
}
```

```{r}
egg_clean <- assign_species_name(egg, species) %>% 
    mutate(volume = 4/3 * pi * length * (width/2)^2) %>% 
    group_by(year, common_name) %>% 
    summarise(mean_volume = mean(volume), .groups = "drop") %>% 
    filter(!is.na(common_name))
```

```{r}
pred_clean <- assign_species_name(pred, species) %>% 
    group_by(year, common_name) %>% 
    summarise(total_count = sum(count), .groups = "drop") %>% 
    filter(!is.na(common_name))
```

```{r, warning = F}
ggplot(egg_clean, aes(x = year, y = mean_volume, color = common_name)) +
    geom_point() +
    geom_line()
```

```{r, warning = F}
ggplot(pred_clean, aes(x = year, y = total_count, fill = common_name)) +
    geom_col()
```


