
<!-- ## Learning Objectives {.unnumbered}

- The advantages of using R packages for organizing code
- Simple techniques for creating R packages
- Approaches to documenting code in packages-->

## R Packages

Most R users are familiar with loading and utilizing packages in their work. And they know how rich CRAN is in providing for many conceivable needs.  Most people have never created a package for their own work, and most think the process is too complicated.  Really it's pretty straighforward and super useful in your personal work.  Creating packages serves two main use cases:

- Mechanism to redistribute reusable code (even if just for yourself)
- Mechanism to reproducibly document analysis and models and their results

Even if you don't plan on writing a package with such broad appeal such as, say, `ggplot2` or `dplyr`, you still might consider creating a package to contain:

- Useful utility functions you write (i.e. a [Personal Package](https://hilaryparker.com/2013/04/03/personal-r-packages/)). Having a place to put these functions makes it much easier to find and use them later.
- A set of shared routines for your lab or research group, making it easier to remain consistent within your team and also to save time.
- The analysis accompanying a thesis or manuscript, making it all that much easier for others to reproduce your results.

::: {.callout-caution icon=false}
### Packages for Creating and Maintaining Packages
The `usethis`, `devtools` and `roxygen2` packages make creating and maintining a package to be a straightforward experience.
:::

### Create a Basic Package

To create a package we're going to use the following packages:

```{r}
#| eval: false

library(devtools)
library(usethis)
library(roxygen2)
```

Thanks to the great [usethis](https://usethis.r-lib.org/) package, it only takes one function call to create the skeleton of an R package using `create_package()`.  Which eliminates pretty much all reasons for procrastination.  To create a package called
`mytools`, all you do is:

```{r}
#| eval: false

usethis::create_package("~/mytools")
```
    
```    
✔ Creating '/home/dolinh/mytools/'
✔ Setting active project to '/home/dolinh/mytools'
✔ Creating 'R/'
✔ Writing 'DESCRIPTION'
Package: mytools
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R (parsed):
    * First Last <first.last@example.com> [aut, cre] (YOUR-ORCID-ID)
Description: What the package does (one paragraph).
License: `use_mit_license()`, `use_gpl3_license()` or friends to
    pick a license
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.3
✔ Writing 'NAMESPACE'
✔ Writing 'mytools.Rproj'
✔ Adding '^mytools\\.Rproj$' to '.Rbuildignore'
✔ Adding '.Rproj.user' to '.gitignore'
✔ Adding '^\\.Rproj\\.user$' to '.Rbuildignore'
✔ Opening '/home/dolinh/mytools/' in new RStudio session
✔ Setting active project to '<no active project>'
```

::: {.callout-caution icon=false}
#### What did the `create_package()` function do?

1. Open a new project called `mytools` (the name of the package) in a new RStduio session.
2. Create a top-level directory structure, including a number of critical files under the [standard R package structure](http://cran.r-project.org/doc/manuals/r-release/R-exts.html#Package-structure):
    a. `DESCRIPTION`file: The most important file, which provides metadata about your package. Edit this file to provide reasonable values for each of the fields, including your  contact information. 
    b. `NAMESPACE` file declares the functions your package exports for external use and the external functions your package imports from other packages.
    c. `R/` directory is where you save all your function scripts and other `.R` files.
    d. `.Rbuildignore` lists files that we need to have around but that should not be included when building the R package from source.
    e. `.Rproj.user` is a directory used internally by RStudio.
3. Add the **Build Tab** to the Environment Pane.
:::

### Add a License

Information about choosing a `LICENSE` is provided in the [Extending R](http://cran.r-project.org/doc/manuals/r-release/R-exts.html#Licensing) documentation.

The `DESCRIPTION` file expects the license to be chose from a predefined list, but
you can use its various utility methods for setting a specific license file, such
as the `MIT` license or the `Apache 2` license:

```{r}
#| eval: false 

usethis::use_apache_license()
```

    ✔ Setting License field in DESCRIPTION to 'Apache License (>= 2.0)'
    ✔ Writing 'LICENSE.md'
    ✔ Adding '^LICENSE\\.md$' to '.Rbuildignore'

Once your license has been chosen, and you've edited your `DESCRIPTION` file with your contact information, a title, and a description, it will look like this:

::: {.callout-tip appearance=minimal}
```
Package: mytools
Title: Halina Do-Linh's Utility R Functions
Version: 0.0.0.9000
Authors@R: 
    person("Halina", "Do-Linh", email = "dolinh@nceas.ucsb.edu", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: A collection of useful R functions that I use for general utilities.
License: Apache License (>= 2)
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.3
```
:::


### Add Code

The skeleton package created contains a directory `R` which should contain your source files.  Add your functions and classes in files to this directory, attempting to choose names that don't conflict with existing packages.  For example, you might add a file `custom_theme` that contains a function `custom_theme()` that you might want to reuse. The `usethis::use_r()` function will help set up you files in the right places.  For example, running:

```{r eval=FALSE}
#| eval: false

usethis::use_r("custom_theme")
```

    ✔ Setting active project to '/home/dolinh/mytools'
    • Modify 'R/custom_theme.R'
    • Call `use_test()` to create a matching test file

creates the file `R/custom_theme` and stores it in the `R` directory, which you can then modify as needed:

```{r}
#| eval: false

custom_theme <- function(base_size = 9) {
    ggplot2::theme(
      axis.ticks       = ggplot2::element_blank(),
      text             = ggplot2::element_text(family = 'Helvetica', 
                                               color = 'gray30', 
                                               size = base_size),
      plot.title       = ggplot2::element_text(size = ggplot2::rel(1.25), 
                                               hjust = 0.5, 
                                               face = 'bold'),
      panel.background = ggplot2::element_blank(),
      legend.position  = 'right',
      panel.border     = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_line(colour = 'grey90', 
                                               linewidth = .25),
      legend.key       = ggplot2::element_rect(colour = NA, 
                                               fill = NA),
      axis.line        = ggplot2::element_blank()
      )
}

```

::: {.callout-caution icon=false}
#### Power of Packages

Remember when we created `custom_theme()` from the [Functions Lesson Section 15.1.4](https://learning.nceas.ucsb.edu/2023-10-coreR/session_15.html#examples-minimizing-work-with-functions)? Now that we've added it to our `mytools` package, we don't have to worry about coyping the code from another file, sourcing the file from another directory, or copying the script from an R Project. 
 
Instead we can leverage the portable functionality of a package to easily access our custom functions **and** maintain the code in one location. 
:::

### Add Dependencies

If your R code depends on functions from another package, you must declare it. In the `Imports` section in the `DESCRIPTION` file, list all the packages your functions depend upon.

In our `custom_theme()` function, we depend on the `ggplot2` package, and so we need to list it as a dependency.

Once again, `usethis` provides a handy helper method:

```{r}
#| eval: false

usethis::use_package("ggplot2")
```

    ✔ Adding 'ggplot2' to Imports field in DESCRIPTION
    • Refer to functions with `ggplot2::fun()`
    
Take a look at the `DESCRIPTION` file again, and you'll see the Imports section has been added, with `ggplot2` underneath.

::: {.callout-tip appearance=minimal}
```
Package: mytools
Title: Halina Do-Linh's Utility R Functions
Version: 0.0.0.9000
Authors@R: 
    person("Halina", "Do-Linh", email = "dolinh@nceas.ucsb.edu", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: A collection of useful R functions that I use for general utilities.
License: Apache License (>= 2)
Encoding: UTF-8
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.2.3
Imports: 
    ggplot2
```
:::

### Add Documentation

Documentation is crucial to add to each of your functions. In the Functions Lesson, we did this using the `roxygen2` package and that same package and approach can be used for packages. 

The `roxygen2` approach allows us to add comments in the source code, where are then converted into Help pages that we can access by typing `?function_name` in the Console.

Let's add documentation for the `custom_theme()` function.

```{r}
#' My custom ggplot theme
#'
#' @param base_size Numeric value of font size of all text elements in plot
#'
#' @return A theme used for ggplot point or line plots
#' @export
#'
#' @examples
#' library(ggplot2)
#' 
#'   ggplot(data = mtcars, aes(x = mpg, y = disp)) +
#'     geom_point() +
#'     custom_theme(base_size = 30)
custom_theme <- function(base_size = 9) {
  ggplot2::theme(
    axis.ticks       = ggplot2::element_blank(),
    text             = ggplot2::element_text(family = 'Helvetica',
                                             color = 'gray30',
                                             size = base_size),
    plot.title       = ggplot2::element_text(size = ggplot2::rel(1.25),
                                             hjust = 0.5,
                                             face = 'bold'),
    panel.background = ggplot2::element_blank(),
    legend.position  = 'right',
    panel.border     = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(colour = 'grey90',
                                             linewidth = .25),
    legend.key       = ggplot2::element_rect(colour = NA,
                                             fill = NA),
    axis.line        = ggplot2::element_blank()
  )
}

```

Once your files are documented, you can then process the documentation using `devtools::document()` to generate the appropriate `.Rd` files that your package needs. The `.Rd` files will appear in the `man/` directory, which is automatically created by `devtools::document()`.

```{r}
#| eval: false

devtools::document()
```

    ℹ Updating mytools documentation
    ℹ Loading mytools
    Writing custom_theme.Rd

We now have a package that we can `check()` and `install()` and `release()`. We'll be using these functions from the `devtools` package, but first let's do some testing.

### Testing

You can test your code using the `testthat` package's testing framework. The `ussethis::use_testthat()` function will set up your package for testing, and then you can use the `use_test()` function to setup individual test files.  For example, in the Functions Lesson we created some tests for our `fahr_to_celsius` functions but ran them line by line in the console.

First, lets add that function to our package. Run the `use_r` function in the console:

```{r}
#| eval: false

usethis::use_r("fahr_to_celsius")
```

Then copy the function and documentation into the R script that opens and save the file.

```{r}
#' Convert temperature data from Fahrenheit to Celsius
#'
#' @param fahr Temperature data in degrees Fahrenheit to be converted
#' @return temperature value in degrees Celsius
#' @keywords conversion
#' @export
#' @examples
#' fahr_to_celsius(32)
#' fahr_to_celsius(c(32, 212, 72))
fahr_to_celsius <- function(fahr) {
  celsius <- (fahr-32)*5/9
  return(celsius)
}
```

Now, set up your package for testing:

```{r}
#| eval: false

usethis::use_testthat()
```
    ✔ Adding 'testthat' to Suggests field in DESCRIPTION
    ✔ Creating 'tests/testthat/'
    ✔ Writing 'tests/testthat.R'


Then write a test for `fahr_to_celsius`:

```{r}
#| eval: false

usethis::use_test("fahr_to_celsius")
```
    ✔ Writing 'tests/testthat/test-fahr_to_celsius.R'
    ● Modify 'tests/testthat/test-fahr_to_celsius.R'

You can now add tests to the `test-fahr_to_celsius.R`, and you can run all of the
tests using `devtools::test()`.  For example, if you add a test to the `test-fahr_to_celsius.R` file:

```{r, eval=FALSE}
#| eval: false

test_that("fahr_to_celsius works", {
  expect_equal(fahr_to_celsius(32), 0)
  expect_equal(fahr_to_celsius(212), 100)
})

```

Then you can run the tests to be sure all of your functions are working using `devtools::test()`:

```{r, eval=FALSE}
#| eval: false

devtools::test()
```

    Loading mytools
    Testing mytools
    ✔ | OK F W S | Context
    ✔ |  2       | test-fahr_to_celsius [0.1 s]
    
    ══ Results ════════════════════════════════════════════════════════════════════════════════════════════════════════════════════════════════════
    Duration: 0.1 s
    
    OK:       2
    Failed:   0
    Warnings: 0
    Skipped:  0

Yay, all tests passed!

### Checking and Installing

Now that your package is built, you can check it for consistency and completeness using `check()`, and then you can install it locally using `install()`, which needs to be run from the parent directory of your module.

```{r, eval=FALSE}
#| eval: false

devtools::check()
devtools::install()
```

Your package is now available for use in your local environment.

### Sharing and Releasing

- **GitHub**: The simplest way to share your package with others is to upload it to a [GitHub repository](https://github.com), which allows others to install your package using the `install_github('mytools','github_username')` function from `devtools`.

- **CRAN**: If your package might be broadly useful, also consider releasing it to CRAN, using the `release()` method from `devtools()`. Releasing a package to CRAN requires a significant amount of work to ensure it follows the standards set by the R community, but it is entirely tractable and a valuable contribution to the science community.  If you are considering releasing a package more broadly, you may find that the supportive community at [ROpenSci](https://ropensci.org) provides incredible help and valuable feeback through their onboarding process.


- **R-Universe**: A newer approach is to link your package release to [R-Universe](https://r-universe.dev), which is an effective way to make it easy to test and maintain packages so that many people can install them using the familiar `install.pacakges()` function in R. In R-Universe, people and organizations can [create their own *universe* of packages](https://github.com/r-universe-org/help#how-to-setup-your-personal-universe), which represent a collection of packages that appear as a CRAN-compatible repository in R. For example, in DataONE we maintaine the DataONE R-Universe (https://dataoneorg.r-universe.dev), which lists the packages we actively maintain as an organization. So, any R-user that wants to install these packages can do so by adding our universe to their list of repositories, and then installing packages as normal. For example, to install the `codyn` package, one could use:

```{r}
#| eval: false

install.packages('codyn', repos = c('https://dataoneorg.r-universe.dev', 'https://cloud.r-project.org'))
```


### Challenge {- .exercise}

Add the other temperature conversion functions with full documentation to your package, write tests to ensure the functions work properly, and then
`document()`, `check()`, and `install()` the new version of the package. Don't forget to update the version number before you install!

### More reading

- Hadley Wickham's awesome book: [R Packages](http://r-pkgs.had.co.nz/)
- Thomas Westlake's blog [Writing an R package from scratch](https://r-mageddon.netlify.com/post/writing-an-r-package-from-scratch/)
- How to set up your [personal R-Universe](https://github.com/r-universe-org/help#how-to-setup-your-personal-universe)



