---
format: html
---

::: callout-tip
## Setup

1.  Open the R Markdown file `collaborator-analysis.Rmd`
    a.  Check the `YAML` and add your name to the `author` field
2.  Load the following libraries at the top of your R Markdown file.

```{r}
#| eval: true
#| code-fold: false
#| message: false
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
```

3.  Read in the dataset and store the data frame as `lobster_traps`

```{r}
#| code-fold: false
lobster_traps <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-sbc.77.8&entityid=66dd61c75bda17c23a3bce458c56ed84")

# lobster_traps <- read_csv("data/Lobster_Trap_Counts_All_Years_20210519.csv")
```

:::

## Convert missing values using `mutate()` and `na_if()`

::: callout-note
### Question 1

The variable `TRAPS` uses -99999 as the code for missing values (see metadata or use `unique()`). This has the potential to cause conflicts with our analyses, so let's convert -99999 to an NA value. Do this using `mutate()` and `na_if()`. Look up the help page to see how to use `na_if()`. 
:::

```{r}
lobster_traps <- lobster_traps %>% 
    mutate(TRAPS = na_if(TRAPS, -99999))
```

## `filter()` practice

::: callout-note
### Question 2
Create and store a subset that does NOT include observations from Naples Reef (`NAPL`). Check your output data frame to ensure that `NAPL` is NOT in the data frame.
:::

```{r}
not_napl <- lobster_traps %>% 
    filter(SITE != "NAPL")
```

::: callout-note
### Question 3
Create and store a subset with lobsters at Carpinteria Reef (`CARP`) OR number of commercial trap floats is greater than 20. Check your output.
:::

```{r}
carp_20_traps <- lobster_traps %>% 
    filter(SITE == "CARP", TRAPS > 20)
```


## `mutate()` practice

::: callout-note
### Question 4
Using the object you created above (we called it `carp_20_traps`), add a new column that contains the proportion of lobster traps at site `CARP`. The total number you are using in your percent calculation is 801. Check your output.
:::

```{r}
# total number of traps at site "CARP" and more than 20 traps counted
lobster_carp_total <- sum(carp_20_traps$TRAPS, na.rm = TRUE)
    
lobster_percent <- carp_20_traps %>% 
  mutate(PERCENT_TRAPS = (TRAPS / lobster_carp_total) * 100)
```

## `group_by() %>% summarize()` practice

::: callout-note
### Question 5
Find the mean using `mean()` of lobster commercial trap floats, grouped by `SITE`. Think about how you want to treat the `NA` values in `TRAPS` (Hint: check the arguments in `mean()`). 
:::

```{r}
mean_lobster_traps <- lobster_traps %>% 
    group_by(SITE) %>%
    summarize(MEAN_TRAPS = mean(TRAPS, na.rm = TRUE))
```

::: callout-note
### Question 6
Find the maximum number of commercial trap floats using `max()` and group by `SITE` and `MONTH`. 
:::

```{r}
max_lobster_traps <- lobster_traps %>% 
    group_by(SITE, MONTH) %>%
    summarize(MAX_TRAPS = max(TRAPS, na.rm = TRUE))
```

