---
bibliography: references.bib
---

## Learning Objectives {.unnumbered}

-   Practice using common cleaning and wrangling functions
-   Practice creating plots using common visualization functions in `ggplot`
-   Practice saving and sharing data visualizations

::: callout-note
## Acknowledgements

These exercises are adapted from Allison Horst's EDS 221: Scientific Programming Essentials Course for the Bren School's Master of Environmental Data Science program.
:::

## About the data {.unnumbered}

These exercises will be using data on abundance, size, and trap counts (fishing pressure) of California spiny lobster (*Panulirus interruptus*) and were collected along the mainland coast of the Santa Barbara Channel by Santa Barbara Coastal LTER researchers [@lter2022].

## Exercise: Collaborate on an analysis and create a report to publish using GitHub Pages

::: callout-tip
### Setup

1.  Create a new repository with a partner
    a.  Determine who is the owner and who is the collaborator
    b.  The owner creates a repository on GitHub titled with both your names (i.e. If Halina and Camila were partners, and Halina is the owner, she would create a repo called `halina-camila`)
        i.  When creating the repository, add a brief description (i.e. coreR R Practice Session: Collaborating on, Wrangling & Visualizing Data), keep the repo Public, and Initialize the repo with a `README` file and an R `.gitignore` template.
    c.  The owner adds the collaborator to the repo
    d.  Both the collaborator and the owner clone the repo into their RStudio

**Step 2 and Step 3 are meant to be completed at the same time. Step 2 is for the collaborator to complete, and Step 3 is for the owner to complete.**

2.  **Collaborator** creates new files for exercise
    a.  The collaborator creates the following R Markdown files that will be used in the exercise:
        i.  Title it: "Owner Analysis" and save it as: `owner-analysis.Rmd`
        ii. Title it: "Collaborator Analysis" and save it as: `collaborator-analysis.Rmd`
        iii. Title it: "Lobster Report" and save it as: `lobster-report.Rmd`
    b.  After creating the files, the collaborator will `stage`, `commit`, write a commit message, and `push` the files to the remote repository (on github)
    c.  The owner pulls the changes and R Markdown files into their local repository (their workspace)
3.  **Owner** downloads data from the EDI Data Portal [SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-sbc.77.8).
    a.  Create two new folders one called `data` and one called `figs`
    b.  Download the following datasets and upload them to the `data` folder:
        i.  Time-series of lobster abundance and size
        ii. Time-series of lobster trap buoy counts
    c.  After creating the `data` folder and adding the datasets, the owner will `stage`, `commit`, write a commit message, and `push` the files to the remote repository (on github)
    d.  The collaborator pulls the changes and datasets into their local repository (their workspace)
:::

### Explore, clean and wrangle data

For this portion of the exercise, the owner will be working with the lobster abundance and size dataset, and the collaborator will be working with the lobster trap buoy counts dataset.

Questions 1-5 you will be working independently since you're working with different datasets, but you're welcome to check in with each other.

::: panel-tabset

#### Owner

{{< include /sections/r-practice-owner-clean-wrangle.qmd >}}

#### Collaborator

{{< include /sections/r-practice-collaborator-clean-wrangle.qmd >}}

::: 

### Create visually appealing and informative data visualization

::: panel-tabset

#### Owner

{{< include /sections/r-practice-owner-visualize.qmd >}}

#### Collaborator

{{< include /sections/r-practice-collaborator-visualize.qmd >}}

:::


### Collaborate on a report and publish using GitHub pages

The final step! Time to work together again. Collaborate with your partner in `lobster-report.Rmd` to create a report to publish to GitHub pages.

::: callout-important
## Code Review
As you're adding owner and collaborator code in the report document, conduct a [lightweight code review](https://en.wikibooks.org/wiki/Introduction_to_Software_Engineering/Quality/Code_Review#:~:text=Code%20review%20practices%20fall%20into,review%20and%20lightweight%20code%20review) and walk each other through your the code for the data visualizations you plan to include in the report (this includes the code you wrote to create the subset for the plot and the code to create the plot). 

This is a lightweight code review, so think brief and give feedback on:

- code readability
- code logic
:::

Make sure your R Markdown is well organized and includes the following elements:

-   citation of the data
-   brief summary of the abstract (i.e. 1-2 sentences)
-   owner analysis and visualizations (you choose which plots you want to include)
    - add alternative text to your plots
-   collaborator analysis and visualizations (you choose which plots you want to include)
    - add alternative text to your plots

Finally, publish on GitHub pages (from owner's repository). Refer back to [Chapter 9](http://localhost:4769/session_12.html) for steps on how to publish using GitHub pages.


## Bonus: Add marine protected area (MPA) designation to the data

The sites `IVEE` and `NAPL` are marine protected areas (MPAs). Add this designation to your data set using a new function called `case_when()`. Then create some new plots using this new variable. Does it change how you think about the data? What new plots or analysis can you do with this new variable? 

::: panel-tabset

### Lobster Abundance & Size Data

Use the object `lobster_abundance` and add a new column called `DESIGNATION` that contains "MPA" if the site is `IVEE` or `NAPL`, and "not MPA" for all other values.

```{r}
lobster_mpa <- lobster_abundance %>% 
    mutate(DESIGNATION = case_when(
    SITE %in% c("IVEE", "NAPL") ~ "MPA",
    SITE %in% c("AQUE", "CARP", "MOHK") ~ "not MPA"
  ))
    
```


### Lobster Trap Buoy Counts Data

Use the object `lobster_traps` and add a new column called `DESIGNATION` that contains "MPA" if the site is `IVEE` or `NAPL`, and "not MPA" for all other values.

```{r}
lobster_traps_mpa <- lobster_traps %>%
    mutate(DESIGNATION = case_when(
    SITE %in% c("IVEE", "NAPL") ~ "MPA",
    SITE %in% c("AQUE", "CARP", "MOHK") ~ "not MPA"
  ))
```

:::



