---
bibliography: references.bib
---

## Learning Objectives {.unnumbered}

-   Practice using common cleaning and wrangling functions
-   Practice creating plots using common visualization functions in `ggplot`
-   Practice saving and sharing data visualizations

::: callout-note
## Acknowledgements

These exercises are adapted from Allison Horst's EDS 221: Scientific Programming Essentials Course for the Bren School's Master of Environmental Data Science program.
:::

## About the data {.unnumbered}

These exercises will be using data on abundance, size and trap counts (fishing pressure) of California spiny lobster (*Panulirus interruptus*) and were collected along the mainland coast of the Santa Barbara Channel by Santa Barbara Coastal LTER researchers [@lter2022].

## Exercise: Collaborate on an analysis and create a report to publish using GitHub Pages

::: callout-tip
## Setup
1. Create a new repository with a partner
    a. Determine who is the owner and who is the collaborator
    b. The owner creates a repository on GitHub titled with both your names (i.e. If Halina and Camila were partners, and Halina is the owner, she would create a repo called `halina-camila`)
        i. When creating the repository, add a brief description (i.e. coreR R Practice Session: Collaborating on, Wrangling & Visualizing Data), keep the repo Public, and Initialize the repo with a `README` file and an R `.gitignore` template.
    c. The owner adds the collaborator to the repo
    d. Both the collaborator and the owner clone the repo into their RStudio
    
**Step 2 and Step 3 are meant to be completed at the same time. Step 2 is for the collaborator to complete, and Step 3 is for the owner to complete.**
    
2. **Collaborator** creates new files for exercise
    a. The collaborator creates the following R Markdown files that will be used in the exercise:
        i. Title it: "Owner Analysis" and save it as: `owner-analysis.Rmd`
        ii. Title it: "Collaborator Analysis" and save it as: `collaborator-analysis.Rmd`
        iii. Title it: "Lobster Report" and save it as: `lobster-report.Rmd`
    b. After creating the files, the collaborator will `stage`, `commit`, write a commit message, and `push` the files to the remote repository (on github)
    c. The owner pulls the changes and R Markdown files into their local repository (their workspace)
    
3. **Owner** downloads data from the EDI Data Portal [SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-sbc.77.8). 
    a. Create two new folders one called `data` and one called `figs`
    b. Download the following datasets and upload them to the `data` folder:
        i. Time-series of lobster abundance and size 
        ii. Time-series of lobster trap buoy counts 
    c. After creating the `data` folder and adding the datasets, the owner will `stage`, `commit`, write a commit message, and `push` the files to the remote repository (on github)
    d. The collaborator pulls the changes and datasets into their local repository (their workspace)
:::

### Explore, clean and wrangle data

::: {.panel-tabset}

#### Owner

```{r}
#| code-fold: false
#| message: false
library(readr)
library(dplyr)
library(here)
library(tidyverse)
library(janitor)
```

#### Collaborator

testing test

```{r}
#| code-fold: false
#| message: false
library(readr)
library(dplyr)
library(here)
library(tidyverse)
library(janitor)
```

:::

### Lobster Abundance and Size Data

### Lobster Trap Buoy Counts Data

```{r}
lobster_abundance <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-sbc.77.8&entityid=f32823fba432f58f66c06b589b7efac6") %>% clean_names()

lobster_fishing_pressure <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-sbc.77.8&entityid=66dd61c75bda17c23a3bce458c56ed84") %>% clean_names()
```

```{r}
lobs_clean <- lobster_fishing_pressure %>%
    mutate(traps = na_if(traps, -99999))
```

```{r}
ggplot(lobs_clean,
       aes(x = fishing_season, y = traps)) +
    geom_col()
```

### Create visually appealing and informative data visualization

::: {.panel-tabset}

#### Owner

```{r}
#| code-fold: false
#| message: false
library(readr)
library(dplyr)
library(here)
library(tidyverse)
library(janitor)
```

#### Collaborator

testing test

```{r}
#| code-fold: false
#| message: false
library(readr)
library(dplyr)
library(here)
library(tidyverse)
library(janitor)
```

:::

### Saving and sharing your work

::: {.panel-tabset}

#### Owner

```{r}
#| code-fold: false
#| message: false
library(readr)
library(dplyr)
library(here)
library(tidyverse)
library(janitor)
```

#### Collaborator

testing test

```{r}
#| code-fold: false
#| message: false
library(readr)
library(dplyr)
library(here)
library(tidyverse)
library(janitor)
```

:::

### Collaborate on a report and publish using GitHub pages

Make sure your R Markdown is well organized and includes the following elements:

-   citation of the data
-   summary of the abstract
-   owner analysis and visualization
-   collaborator analysis and visualization
-   code review!
-   publish on github pages (from owner's repository)
