## Learning Objectives {.unnumbered}

- Practice cleaning and wrangling spatial data
- Practice joining spatial data
- Practice visualizing spatial data and making maps

::: callout-note
These exercises are adapted from the [`lterdatasampler` package](https://lter.github.io/lterdatasampler/index.html).
:::

## About the data
These exercises will be using pika stress metrics and Colorado Rockies elevation data collected by researchers at the Niwot Ridge LTER research site. Data will be accessed from the [EDI Portal Physiological stress of American pika (*Ochotona princeps*) and associated habitat characteristics for Niwot Ridge, 2018 - 2019](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-nwt.268.1). 

## Exercise: 

::: callout-tip
## Setup

0. Make sure you're in the right project (`training_{USERNAME}`) and use the `Git` workflow by `Pull`ing to check for any changes in the remote repository (aka repository on GitHub).

1. Create a new R Markdown file.
    a. Title it “R Practice: Spatial Data and Making Maps”.
    b. Save the file and name it “r-practice-spatial-data-maps”.
    
2. Load the following libraries at the top of your R Markdown file.

```{r}
#| warning: false
#| message: false
library(readr)
library(tidyr)
library(dplyr)
library(sf)
library(scales)
library(ggplot2)
library(leaflet)
library(stringr)
```

3. Obtain pika and elevation data from the [EDI Portal Physiological stress of American pika (Ochotona princeps) and associated habitat characteristics for Niwot Ridge, 2018 - 2019](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-nwt.268.1). Download the following files:
    - `pika_fecal_glut.aw.data.csv` (Fecal glucocorticoid concentration)
    - `pika_fecal_glut_habitat.aw.data.csv` (Habitat characteristics)
    
**Note:** It’s up to you on how you want to download and load the data! You can either use the download links (obtain by right-clicking the “Download” button and select “Copy Link Address” for each data entity) or manually download the data and then upload the files to RStudio server.

4. Organize your R Markdown in a meaningful way. This is up to you! Consider the different ways we’ve organized previous files using: headers, bold text, naming code chunks, comments in code chunks. What is most important is organizing and documenting the file so that your future self (or if you share this file with others!) understands it as well as your current self does right now.

5. Use the `Git` workflow. After you've set up your project and uploaded your data go through the workflow: ```Stage > Commit > Pull > Push```
    - **Note:** You also want to `Pull` when you first open a project.
:::

## Exercise: Practice Cleaning & Wrangling data

### Read in the data

::: callout-note
#### Question 1

Read in the data and store the data frame `pika_fecal_glut.aw.data.csv` (Fecal glucocorticoid concentration) to an object called `pika`, and store the second data frame `pika_fecal_glut_habitat.aw.data.csv` (Habitat characteristics) to an object called `elevation`
:::

```{r}
# read in data using download links
pika <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-nwt.268.1&entityid=43270add3532c7f3716404576cfb3f2c")

elevation <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-nwt.268.1&entityid=6a10b35988119d0462837f9bfa31dd2f")

# read in data from the data directory after manually downloading data 
pika <- read_csv("data/pika_fecal_glut.aw.data.csv")
elevation <- read_csv("data/pika_fecal_glut_habitat.aw.data.csv")
```

### Get familiar with the data

::: callout-note
#### Question 2

Always, always look at your data. Use functions to explore the data you just read in. Think about which functions we've been using to see the data, access information about columns, unique observations, data types and structure of the data frame.

**Note:** We will only use the columns `Station` and `Elev_M` from the `elevation` table, so spend more time explore the `pika` data. Also, some quick metadata: `Concentration_pg_g` in the `pika` table is the stress measurement and was taken from fecal pellets.
:::

```{r}
# view data in a spreadsheet-style viewer
View(pika)

# returns dimensions of the dataframe by number of rows and number of cols
dim(pika)

# returns the top six rows of the dataframe
head(pika)

# returns all the columns and some info about the cols
glimpse(pika)

# similar to glimpse but returns some summary statistics about the cols
summary(pika)

# returns column names 
names(pika)

# returns unique values in a column. In this case we can see all the different bird species IDs
unique(pika$Site)
```


### Clean & Wrangle `pika` and `elevation` tables

We are going to be using some of the cleaning steps that the `lterdatasampler` developers used. Some of their cleaning steps may be beyond what we've covered in this course, but it's overall straightforward and we'll be guiding you through every step. 

Let's start with the `pika` data.

::: column-tip
#### Question 3

From the `pika` data we only want the columns:

- `Date`
- `Station`
- `UTM_Easting`
- `UTM_Northing`
- `Concentration_pg_g`
- `Site`
- `Sex`

Write code to subset the data so it only contains those columns and save it to an object called `pika_clean`.
:::

```{r}
pika_clean <- pika %>% 
    select(-Notes,
           -Vial,
           -Plate,
           -Biweek)
```

::: column-tip
#### Question 4

Next, we want to convert data types. 

The variables `Station`, `Site`, and `Sex` all have a class of `character` (you can find out what data type a variable is by using the function `class()` i.e. `class(pika_clean$Station)`). We want to convert these three variables so that they have a class of `factor`. Do this within a `mutate()` and use the function `as.factor()`. Don't forget you can always check the help page using `?`.

In the rest of these cleaning steps, we'll be overriding `pika_clean`, meaning you don't need to create a new object.

Check that the conversation worked using `class()`.
:::

```{r}
# wait to convert Date
pika_clean <- pika_clean %>% 
    mutate(
        Station = as.factor(Station),
        Site = as.factor(Site),
        Sex = as.factor(Sex)
    )
```


::: column-tip
#### Question 5

Next, we want to spell out the abbreviations in `Site`. With `unique()`, we find that `Site` has three unique site IDs: `CG`, `WK`, `LL`. And from the metadata, we know that:

- CG = Cable Gate
- WK = West Knoll
- LL = Long Lake

Write code to spell out the abbreviations within a `mutate()` using the function `recode()`. `recode()` is an extremely useful function that allows you to replace values within a vector. View the help page or use Google.

Check that your code worked using `unique()`.
:::

```{r}
pika_clean <- pika_clean %>% 
    mutate(Site = recode(
        Site,
        "CG" = "Cable Gate",
        "WK" = "West Knoll",
        "LL" = "Long Lake"
    ))
```


::: callout-tip
#### Question 6

We're almost done! Take a look at the `pika_clean` data using `View()`. The table might be better organized if we had the variables `Station` and `Site` next to each other. We can change that using the function `relocate()`. `relocate()` is another useful function that allows us to change the column order using the arguments `.before` and `.after`. If needed, view the help page or use Google.

Let's change the order of the columns so that `Site` is before `Station`, and `Sex` is after `UTM-Northing`. This does not need to be within a `mutate()` because we are not changing any of the values within a column / variable.

Check that your code worked by viewing `pika_clean` using `View()`.
:::

```{r}
pika_clean <- pika_clean %>% 
    relocate(Site, .before = Station) %>% 
    relocate(Sex, .after = UTM_Northing)
```

We're done! These are the all the cleaning and wrangling steps we are going to do for `pika`. We completed these steps one by one in their own code chunks, but know that you could have piped into each step in a single code chunk. See the code chunk below:

```{r}
# cleaning and wrangling using a piped sequence
pika_clean <- pika %>% 
    # Question 3
    select(-Notes, 
           -Vial, 
           -Plate, 
           -Biweek) %>% 
    mutate(
        # Question 4
        Station = as.factor(Station),
        Site = as.factor(Site),
        Sex = as.factor(Sex),
        # Question 5 (Note: within the same mutate)
        Site = recode(
            Site,
            "CG" = "Cable Gate",
            "WK" = "West Knoll",
            "LL" = "Long Lake"
        )
    ) %>%
    # Question 6
    relocate(Site, .before = Station) %>%
    relocate(Sex, .after = UTM_Northing)
```


Now let's clean and wrangle `elevation`.

::: column-tip
#### Question 7

From the `elevation` data we only want the columns:

- `Station`
- `Elev_M`

Write code to subset the data so it only contains those columns and save it to an object called `elevation_clean`.
:::

```{r}
elevation_clean <- elevation %>% 
    select(Station,
           Elev_M)
```

::: column-tip
#### Question 8

Lastly, we want to convert data types. 

The variable `Station` has a class of `character` (you can find out what data type a variable is by using the function `class()` i.e. `class(elevation_clean$Station)`). We want to convert this variable so that it has a class of `factor`. Do this within a `mutate()` and use the function `as.factor()`. Don't forget you can always check the help page using `?`.

In the rest of these cleaning steps, we'll be overriding `elevation_clean`, meaning you don't need to create a new object.

Check that the conversation worked using `class()`.
:::

```{r}
elevation_clean <- elevation %>% 
    mutate(
        Station = as.factor(Station)
    )
```

If we combined these cleaning and wrangling steps in a piped sequence it would look like this:

```{r}
# cleaning and wrangling using a piped sequence
elevation_clean <- elevation %>% 
    select(Station,
           Elev_M) %>%
    mutate(Station = as.factor(Station))
```

### Join `pika_clean` and `elevation_clean`



```{r}
# using pipe
pikas <- pika_clean %>% 
    full_join(y = elevation_clean,
              by = "Station")

# without pipe
pikas <- full_join(x = pika_clean,
                   y = elevation_clean,
                   by = "Station")
```









