---
format: html
---

NOTES:
- add a ggsave step
- add instructions: FIRST you're going to create the subsets you need to create the visualizations, THEN you're going to make the graphs pretty
- code fold graph outputs; add text "your graph should look something like this"

::: callout-note
### Question 7
Create a multi-panel plot of lobster carapace length (`SIZE_MM`) using `ggplot()`, `geom_histogram()`, and `facet_wrap()`. Use the variable `SITE` in `facet_wrap()`.
:::

```{r}
#| eval: true
#| message: false
#| warning: false
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)


lobster_traps <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-sbc.77.8&entityid=66dd61c75bda17c23a3bce458c56ed84") %>% 
    mutate(TRAPS = na_if(TRAPS, -99999))

ggplot(data = lobster_traps, aes(x = TRAPS)) +
    geom_histogram() +
    facet_wrap(~SITE)
```


::: callout-note
### Question 8
Create a bar graph that shows the amount of small and large sized carapace lobsters at each `SITE` from 2019-2021.
:::

First, you'll need to create a new dataset subset:

- `filter()` for the years 2019, 2020, and 2021
- Add a new column called `SIZE_BIN` that contains the values "small" or "large".  A "small" carapace size is <= 70 mm, and a "large" carapace size is greater than 70 mm. Use `mutate()` and `if_else()`. Check your output
- Remove the `NA` values from the subsetted data. Hint: check out `drop_na()`. Check your output

```{r}
#| eval: true
lobster_traps_5yrs <- lobster_traps %>% 
    filter(YEAR %in% c(2017, 2018, 2019, 2020, 2021)) %>% 
    group_by(FISHING_SEASON) %>% 
    summarize(total_traps = sum(TRAPS))
```

Next, create a bar graph using `ggplot()` and `geom_bar()`. Note that `geom_bar()` automatically creates a stacked bar chart. Try using the argument `position = "dodge"` to make the bars side by side. Pick which bar position you like best.

```{r}
#| eval: true
# total number of traps at each site from the past three years
ggplot(data = lobster_traps_5yrs, aes(x = FISHING_SEASON, y = total_traps)) +
    geom_bar(stat = "identity")
```

::: callout-note
### Question 9

Create a ggplot graph of the number of total lobsters observed (y-axis) by year (x-axis) in the study, grouped by `SITE`
:::

First, you'll need to create a new dataset subset:

- Group the data by `SITE` AND `YEAR`
- Calculate 


## Create a summary table that finds the total counts (see: n()), and mean carapace lengths of lobsters observed in the dataset by site and year.

```{r}
#| eval: true
lobsters_traps_summarize <- lobster_traps %>% 
  group_by(SITE, YEAR) %>% 
  summarize(count = n()) %>% 
  ungroup()
```


```{r}
#| eval: true
ggplot(data = lobsters_traps_summarize, aes(x = YEAR, y = count)) +
  geom_point(aes(color = SITE)) +
  geom_line(aes(color = SITE)) 
```

## Bonus

Take the subset `lobster_size` and add a new column called `DESIGNATION` that contains “MPA” if the site is “IVEE” or “NAPL”, and “not MPA” for all other outcomes.

```{r}
lobster_traps_mpa <- lobster_traps %>% 
    mutate(DESIGNATION = if_else(SITE %in% c("IVEE", "NAPL"), true = "MPA", false = "not MPA")) %>% 
    select(c("YEAR", "FISHING_SEASON", "SITE", "TRAPS", "DESIGNATION")) %>% 
    filter(YEAR %in% c(2019, 2020, 2021))
```

What mpa has the largest

```{r}
ggplot(data = lobster_traps_mpa, aes(x = YEAR, Y = TRAPS, fill = DESIGNATION)) +
    geom_bar(position = "dodge") +
    facet_wrap(~SITE)
```

