---
title: "Data Packaging"
author: "Bryce Mecum <mecum@nceas.ucsb.edu>"
date: "7/31/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This document briefly covers two common tasks:

- Uploading a data file to the Arctic Data Center
- Uploading an entire Data Package to the Arctic Data Center

The R `dataone` package offers some convenient ways to do this, particularly some custom R classes which encapsulate a lot of the complexity of Objects and Data Packages.

First, let's load our usual libraries and connect to our Member Node:

```{r}
library(dataone)
library(datapack)
```

Here we use a `D1Client` which is another way of connecting to a Member Node:

```{r}
client <- D1Client("STAGING", "urn:node:mnTestARCTIC")
```

Alternatively, we could also create an `MNode` object directly.

## Upload a single data file

First let's make an example data file (CSV in this case) to upload:

```{r}
data("co2") # Built-in dataset in R
write.csv(co2, "./co2.csv")
```

Every Object we upload then just needs a `DataObject` class instance for it:

```{r}
my_object <- new("DataObject", 
                 filename = "./co2.csv",
                 format = "text/csv")

my_object <- setPublicAccess(my_object) # Make it public readable!
```

And then we just need to upload it:

```{r}
uploadDataObject(client, my_object)
```

## Upload a simple Data Package

The steps to upload an entire package aren't that much more complex.
First, let's create an example EML file:

```{r}
library(EML)

title <- "Test dataset to show submitting via R client"
me <- as.person("Bryce Mecum <mecum@nceas.ucsb.edu>")
dataset <- new("dataset",
               title = title,
               creator = me,
               contact = me)

eml_pid <- paste0("urn:uuid", uuid::UUIDgenerate())

eml <- new("eml",
           packageId = eml_pid,
           system = "uuid",
           dataset = dataset)

eml_path <- "~/my_eml.xml"
write_eml(eml, eml_path)
eml_validate(eml_path)
```

And then we just use the `DataPackage` class:

```{r}
my_package <- new("DataPackage")

my_metadata <- new("DataObject", 
                   format = "eml://ecoinformatics.org/eml-2.1.1", 
                   filename = eml_path)

my_object <- setPublicAccess(my_object) # Make it public readable!

my_object <- new("DataObject", 
                 filename = "./co2.csv",
                 format = "text/csv")

my_object <- setPublicAccess(my_object) # Make it public readable!

addData(my_package, my_metadata)
addData(my_package, my_object, mo = my_metadata)
```

```{r}
uploadDataPackage(client, my_package, public = TRUE)
```
