# Reproducibility and Provenance

## Learning Objectives

In this lesson, you will learn:

- About the importance of computational reproducibility
- The role of provenance metadata
- Tools and techniques for reproducibility supportred by the Arctic Data Center
- How to build a reproducible paper in RMarkdown

A great overview of this approach to reproducible papers comes from:

- Ben Marwick, Carl Boettiger & Lincoln Mullen (2018) **Packaging Data Analytical Work Reproducibly Using R (and Friends)**, The American Statistician, 72:1, 80-88, [doi:10.1080/00031305.2017.1375986](https://doi.org/10.1080/00031305.2017.1375986)

This lesson will draw from existing materials:

- [Accelerating synthesis science through reproducible science practices](files/2019-10-11-repro-sci.pdf)
- [rrtools](https://github.com/benmarwick/rrtools)
- [Reproducible papers with RMarkdown](https://nceas.github.io/oss-lessons/reproducible-papers-with-rmd/reproducible-papers-with-rmd.html)

To start a reproducible paper with `rrtools`, run:

```{r eval=FALSE}
devtools::install_github("benmarwick/rrtools")
setwd("..")
rrtools::use_compendium("arcticpaper")
```

Then, add some more structure to the package:

```{r eval=FALSE}
usethis::use_apl2_license(name="Matthew B. Jones")
rrtools::use_readme_rmd()
rrtools::use_analysis()
```

Now write a reproducible paper!
