--- 
title: "Open Science Synthesis for the Delta Science Program: Week 3"
date: "November 1-5, 2021"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: ""
always_allow_html: yes
---

# Open Science Synthesis for the Delta Science Program: Week 3 {-#reproducible-research-techniques-for-synthesis}

Placeholder


## Schedule
### Code of Conduct
#### The Power of Open
### About this book

<!--chapter:end:index.Rmd-->

# Session 1: Time Series and Forecasting

```{r, child = 'sections/time-series-forecasting.Rmd'}
```

<!--chapter:end:session_01.Rmd-->

# Session 2: Collaborative synthesis

<!--chapter:end:session_02.Rmd-->

# Session 3: Web based data archival

```{r, child = 'sections/metadata-data-best-practices-intro.Rmd'}
```

```{r, child = 'sections/metadata-knb-data-documentation-hands-on.Rmd'}
```

<!--chapter:end:session_03.Rmd-->

# Session 4: Programmatic metadata and data access

```{r, child = 'sections/data-identifers-contentid.Rmd'}
```

```{r, child = 'sections/metadata-programming-publishing.Rmd'}
```

<!--chapter:end:session_04.Rmd-->

# Session 5: Model Analysis

```{r, child = 'sections/model-analysis.Rmd'}
```



<!--chapter:end:session_05.Rmd-->

# Session 6: Shiny

```{r, child = 'sections/visualisation-shiny.Rmd'}
```

<!--chapter:end:session_06.Rmd-->

# Session 7: Collaborative synthesis

<!--chapter:end:session_07.Rmd-->

# Session 8: Reproducible workflows revisited

```{r, child = 'sections/parallel-computing-in-r.Rmd'}
```

```{r, child = 'sections/Workflows-and-continuous-integration.Rmd'}
```

<!--chapter:end:session_08.Rmd-->

# Session 9: Communicating Your Research: The Message Box

```{r, child = 'sections/communication.Rmd'}
```

<!--chapter:end:session_09.Rmd-->

# Session 10: Collaborative synthesis

<!--chapter:end:session_10.Rmd-->

# Session 11: Collaborative Synthesis

<!--chapter:end:session_11.Rmd-->

# Session 12: Collaborative Synthesis

<!--chapter:end:session_12.Rmd-->

# Session 13: Collaborative Synthesis and Wrap-up



<!--chapter:end:session_13.Rmd-->

