# Publication Graphics

## Learning Objectives

In this lesson, you will learn:

- The basics of the `ggplot2` package to create static plots
- How to use `ggplot2`'s theming abilities to create publication-grade graphics
- The basics of the `leaflet` package to create interactive maps

## Overview

ggplot2 is a popular package for visualizing data in R.
From the [home page](http://ggplot2.tidyverse.org/):

> ggplot2 is a system for declaratively creating graphics, based on The Grammar of Graphics. You provide the data, tell ggplot2 how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.
It's been around for years and has pretty good documentation and tons of example code around the web (like on [StackOverflow](https://stackoverflow.com/questions/tagged/ggplot2)).
This lesson will introduce you to the basic components of working with ggplot2.

### ggplot vs base vs lattice vs XYZ...

R provides **many** ways to get your data into a plot.
Three common ones are,

- "base graphics" (`plot()`, `hist()`, etc`)
- lattice
- ggplot2
All of them work!
I use base graphics for simple, quick and dirty plots.
I use ggplot2 for most everything else.
ggplot2 excels at making complicated plots easy and easy plots simple enough.


## Setup

First, let's load the packages we'll need:

```{r load_packages, message = F, warning = F}
  library(leaflet)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(DT)
  library(sf)
  library(ggmap) # devtools::install_github("dkahle/ggmap")
```



### Load salmon escapement data

You can load the data table directly from the KNB Data Repository, if it isn't already present on your local computer.  This technique only downloads the file if you need it.

```{r read_data}
data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e"

esc <- tryCatch(
    read.csv("data/escapement.csv", stringsAsFactors = FALSE),
    error=function(cond) {
        message(paste("Escapement file does not seem to exist, so get it from the KNB."))
        esc <- read.csv(url(data_url, method = "libcurl"), stringsAsFactors = FALSE)
        return(esc)
    }
)

head(esc)
```

Now that we have the data loaded, let's calculate annual escapement by species and region:

```{r mean_esc}
annual_esc <- esc %>% 
  separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  group_by(Species, SASAP.Region, Year) %>% 
  summarize(escapement = sum(DailyCount)) %>% 
  filter(Species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink"))

head(annual_esc)
```

That command used a lot of the dplyr commands that we've used, and some that are new. The `separate` function is used to divide the sampleDate column up into Year, Month, and Day columns, and then we use `group_by` to indicate that we want to calculate our results for the unique combinations of species, region, and year.  We next use `summarize` to calculate an escapement value for each of these groups. Finally, we use a filter and the `%in%` operator to select only the salmon species.


## ggplot

Every graphic you make in ggplot2 will have at least one aesthetic and at least one geom (layer). The aesthetic maps your data to your geometry (layer). Your geometry specifies the type of plot we're making (point, bar, etc.).

Now, let's plot our results using `ggplot`. `ggplot` uses a mapping *aesthetic* (set using `aes()`) and a *geometry* to create your plot. Additional geometries/aesthetics and theme elements can be added to a `ggplot` object using `+`.

```{r plot_esc}
ggplot(annual_esc, aes(x = Species, y = escapement)) +
  geom_col()
```

What if we want our bars to be blue instad of gray? You might think we could run this:

```{r}
ggplot(annual_esc, aes(x = Species, y = escapement, fill = "blue")) +
  geom_col()
```

Why did that happen?

Notice that we tried to set the fill color of the plot inside the mapping aesthetic call. What we have done, behind the scenes, is create a column filled with the word "blue" in our dataframe, and then mapped it to the `fill` aesthetic, which then chose the default fill color of red. 

What we really wanted to do was just change the color of the bars. If we want do do that, we can call the color option in the `geom_bar` function, **outside** of the mapping aesthetics function call.

```{r}
ggplot(annual_esc, aes(x = Species, y = escapement)) +
  geom_col(fill = "blue")
```

What if we did want to map the color of the bars to a variable, such as region.

`ggplot` is really powerful because we can easily get this plot to visualize more aspects of our data.

```{r}
ggplot(annual_esc, aes(x = Species, y = escapement, fill = SASAP.Region)) +
  geom_col()
```

Here is an example of a different aesthetic and geometry mapping we can create with this data. You can also add a title, adjust labels, and include a built in theme.

```{r}
ggplot(filter(annual_esc, SASAP.Region == "Kodiak"), aes(x = Year, y = escapement, color = Species)) + 
    geom_line() +
    geom_point() +
    ylab("Escapement") +
    ggtitle("Kodiak Salmon Escapement") +
    theme_bw()

```

Note how I added a filter call in my `ggplot` function call to only plot data from Kodiak. You can easily create plots like this for every region using `facet_wrap`

```{r}
ggplot(annual_esc, aes(x = Year, y = escapement, color = Species)) + 
    geom_line() +
    geom_point() +
    facet_wrap(~SASAP.Region, scales = "free_y") +
    ylab("Escapement") +
    theme_bw()

```

## Leaflet

Now let's convert the escapement data into a table of just the unique locations:

```{r uniq_locations}
locations <- esc %>% 
  distinct(Location, Latitude, Longitude) %>% 
  drop_na()
```

And display it as an interactive table:

```{r display_locations}
datatable(locations)
```

Then making a leaflet map is (generally) only a couple of lines of code:

```{r simple_map}
leaflet(locations) %>% 
  addTiles() %>% 
  addMarkers(~ Longitude, ~ Latitude, popup = ~ Location)
```

The `addTiles()` function gets a base layer of tiles from [OpenStreetMap](https://www.openstreetmap.org/) which is an open alternative to Google Maps.
`addMarkers` use a bit of an odd syntax in that it looks kind of like `ggplot2` code but uses `~` before the column names.
This is similar to how the `lm` function (and others) work but you'll have to make sure you type the `~` for your map to work.

You can also use `leaflet` to import Web Map Service (WMS) tiles. Here is an example that utilizes the General Bathymetric Map of the Oceans [(GEBCO)](https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/#getcapabilities) WMS tiles. In this example, we also demonstrate how to create a more simple circle marker.

```{r}
leaflet(locations) %>% 
  addWMSTiles("https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/mapserv?",
              layers = 'GEBCO_LATEST',
              attribution = "Imagery reproduced from the GEBCO_2014 Grid, version 20150318, www.gebco.net") %>%
  addCircleMarkers(lng = ~Longitude,
                   lat = ~Latitude,
                   popup = ~ Location,
                   radius = 5,
                   # set fill properties
                   fillColor = "salmon",
                   fillOpacity = 1,
                   # set stroke properties
                   stroke = T,
                   weight = 0.5,
                   color = "white",
                   opacity = 1)
```

Leaflet has a ton of functionality that can enable you to create some beautiful, functional maps with relative ease. [Here](https://pages.github.nceas.ucsb.edu/NCEAS/sasap-data/language_vis.html) is an example of some we created as part of the SASAP project, created using the same tools we showed you here. This map hopefully gives you an idea of how powerful the combination of RMarkdown and GitHub pages can be. 


## Static maps with geom_sf

We can also create a similar static map using `ggplot` with the `geom_sf` geometry. To
do so, first we need to transform our locations data frame into an `sf` object by
providing the coordinate reference system (EPSG:4326 is the CRS for geocoordinates in WGS84).
We also transform the points to `EPSG:3857` which is the CRS used for rendering 
maps in Google Maps, Stamen, and OpenStreetMap, among others.

```{r}
locations_sf <- locations %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
locations_sf_3857 <- st_transform(locations_sf, 3857)
```

Next, let's grab a base map from the Stamen map tile server covering the region of interest.
This includes a fix to transform the bounding box (which starts in `EPSSG:326`) to also be in the `EPSG:3857` CRS,
which is the projection that the map raster is returned in from Stamen.


```{r message=FALSE}
# Define a function to fix the bbox to be in EPSG:3857
# See https://github.com/dkahle/ggmap/issues/160#issuecomment-397055208
ggmap_bbox_to_3857 <- function(map) {
  if (!inherits(map, "ggmap")) stop("map must be a ggmap object")
  # Extract the bounding box (in lat/lon) from the ggmap to a numeric vector, 
  # and set the names to what sf::st_bbox expects:
  map_bbox <- setNames(unlist(attr(map, "bb")), 
                       c("ymin", "xmin", "ymax", "xmax"))
  
  # Coonvert the bbox to an sf polygon, transform it to 3857, 
  # and convert back to a bbox (convoluted, but it works)
  bbox_3857 <- st_bbox(st_transform(st_as_sfc(st_bbox(map_bbox, crs = 4326)), 3857))
  
  # Overwrite the bbox of the ggmap object with the transformed coordinates 
  attr(map, "bb")$ll.lat <- bbox_3857["ymin"]
  attr(map, "bb")$ll.lon <- bbox_3857["xmin"]
  attr(map, "bb")$ur.lat <- bbox_3857["ymax"]
  attr(map, "bb")$ur.lon <- bbox_3857["xmax"]
  map
}

bbox <- c(-170, 52, -130, 64)   # This is roughly southern Alaska
ak_map <- get_stamenmap(bbox, zoom = 4)
ak_map_3857 <- ggmap_bbox_to_3857(ak_map)
```

 Finally, plot both the base raster map with the points overlayed, which is easy now
 that everything is in the same projection (3857):

```{r message=FALSE}
ggmap(ak_map_3857) + 
  geom_sf(data = locations_sf_3857, inherit.aes = FALSE)
```

## Resources

- Lisa Charlotte Rost. (2018) [Why not to use two axes, and what to use instead: The case against dual axis charts](https://blog.datawrapper.de/dualaxis/)
